### 2010-12-29 Documentation/atropos/atropos.tex
ENH: submitted version of atropos paper.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@733 fd2da0a4-d968-48df-9cb5-d8c496287cbf

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$
(cf. Eqn. (\ref{eq:likelihood})), 
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions,
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}.
\end{align}
The $t_{ik}$ is the prior probability value at site $i$ which was
mapped, typically by image registration, to the local image
from a template data set.  The user may also
choose mixing proportions equal to
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
\end{align}
via the command line interface to the posterior formulation.

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alterative to Boykov's solution.
Furthermore, one may automate the
initialization process.

@@

EM will iterate toward a local maximum.  We track
convergence by summing up the maximum posterior
probability at each site over the segmentation domain.  

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
Initialization provides an estimate before the iterative optimization in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}
\subsection{The Atropos User Interface}
As with other classes that comprise ANTs, Atropos uses the Insight
Toolkit as a developmental foundation.  This allows us to take
advantage of the mature portions of ITK (e.g. image IO) and ensures
the integrity of the ancillary processes such as those facilitated by
the underlying statistical framework.  Although Atropos is publicly
distributed with the rest of the ANTs package, we plan to contribute
the Atropos to the Insight Toolkit where it can be vetted and further
improved by other interested researchers.

An overview of the core components of Atropos can be gleaned, in part, from the flowchart depicted in 
Fig.~\ref{fig:flowchart}.  To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.  
The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.  Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.  Here we  describe these options from the unique perspective of  implementation.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing
\subsubsection{Integrating N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight Toolkit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.

@@

data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We

@@

of bias on our N4 and Atropos and also the validity of our

@@

Dice overlap, are shown in figure~\ref{fig:bweb1}  Because overlap

@@

is in figure~\ref{fig:bwebrf40}.  We supply the information necessary

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.  The
  legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4 assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=1$, the priors are only maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

@@

We first convert each of the 69 labels within the original evaluation
dataset to an individual image.  The remaining steps,
summarized briefly, are the same for each of the 19 subjects.  We
select one subject as an unlabeled target.  The other 18 datasets are
then mapped (as in the script above) to that subject.  We then deform,
individually, the 69 $\times$ 18 label images to the unlabeled
subject.  The label probability map is gained by averaging the 18
deformed images associated with each label.  We repeat this for each
subject.  The following parameters are the most relevant to this discussion:
{\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
-m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
Dice overlap, are shown in figure~\ref{fig:hammers}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small The figure compares the Dice
overlap results from Atropos versus the raw results from majority
voting for each of 68 neuroanatomical regions and, in addition, the
unlabeled portions of the brain from the Hammers evaluation dataset.
We evaluated Atropos via N-fold cross-validation and employed
PriorProbabilityImages for each class where probabilities are gained
by averaging mapped subject labels. The color coding highlights those
regions that have the highest (yellow) and lowest (pink) improvement.
The significance of the improvement, measured by pairwise T-test, is
also shown as is a trinary coding of that improvement as: $+$
significant improvement, $-$ performance reduction, $\sim$ no change.}
\label{fig:hammers}
\end{center}
\end{figure}

@@

\section{Discussion}
We introduced Atropos, the theory and implementation
details and documented its performance in a variety of use cases.  We
also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
Furthermore, we used multiple subject BrainWeb data to build dataset
specific priors that provided the most consistent segmentation
performance across tissues.  Finally, we used majority voting to
initialize an Atropos EM solution to a 69 class brain parcellation
problem.  Significant improvements were gained in multiple brain
regions, in particular in temporal lobe cortex, the hippocampi and
amygdalae and the lateral ventricles.  This work, in summary, proves
the applicability of Atropos in both basic and extended use cases.

\subsection{Performance on BrainWeb Data}
Atropos results are competitive with the state of the art.  For
instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
noise BrainWeb single subject data finding 0.932 (GM) and .961 (WM)
Dice. Results on 40\% RF were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
insensitivity to bias similar to our own best results on the 40\% RF, 3\% 
noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice for GM is
0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
results (BrainWeb single 3\% noise) of 0.962 (0\%cRF), 0.964 (20\% RF)
and 0.956 (40\%RF) which are slightly higher than either SPM5 or Atropos
results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
for comparison.  Topology preserving methods also perform well.
\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF BrainWeb
single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
respectively.  These are excellent numbers given the additional
constraint applied to the segmentation.  
\cite{Bazin2007c} proposed TOADS and, estimating from the
paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
Perhaps the most recent balanced evaluation was performed in
\citep{Klauschen2009}, which reports confusion matrix numbers, rather
than Dice overlap.  Because the absolute true number of GM and WM
voxels for BrainWeb are known, we can convert the confusion matrix to
Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is 0.8847
and 0.9085, respectively, while FreeSurfer and FSL's accuracy is
lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;
best WM Dice is from FSL: 0.9499.  We note that \cite{Klauschen2009}
used a comprehensive evaluation where quality of brain extraction also
contributed to the outcome.  Thus, the results must be interpreted
slightly differently than those from other papers.

\subsection{Performance on Hammers Data}
Our prior work, \citep{Avants2011}, showed that the majority vote
initialization provided to Atropos by ANTs template mapping is
competitive with \cite{Heckemann2006}.  The Atropos EM extension of
improved these results further.  However, the Atropos EM segmentation
performed significantly worse.  This is not surprising, in that
Atropos EM assumes that signal from the likelihood and MRF term is
valuable in improving the segmentation.  This assumption held for
amygdala and lateral ventricles among other areas.  However, in
pallidum and corpus callosum (the most significant areas with loss of
performance), this is not true.  We believe the explanation is that
the intensity varies within these structures and that a more complex
intensity model (or finer parcellation) would be needed here.  An
alternative solution would be to use boundary conditions for these
structures, as in the {\ttfamily PriorLabelImage} Atropos initialization option.

>>>>

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$
(cf. Eqn. (\ref{eq:likelihood})), 
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions,
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}.
\end{align}
The $t_{ik}$ is the prior probability value at site $i$ which was
mapped, typically by image registration, to the local image
from a template data set.  The user may also
choose mixing proportions equal to
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}},
\end{align}
via the command line interface to the posterior formulation.

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alterative to Boykov's solution.
Furthermore, one may automate the initialization process.  We compare
our sparse prior labeling strategy to fully automated methods in the
evaluation section.

@@

EM will iterate toward a local maximum.  We track convergence by
summing up the maximum posterior probability at each site over the
segmentation domain.  The E-step, above, depends upon the selected
coding strategy \citep{Besag1986}.  Atropos may use either a
classical, sequential checkerboard update or a synchronous update of
the labels, the latter of which is commonly used in practice.
Synchronous update does not guarantee convergence but we employ it by
default due to its intrinsic parallelism and speed.  The user may
alternatively select checkerboard update if he or she desires
theoretical convergence guarantees.  However, we have not identified
performance differences, relative to ground truth, that convince us of
the absolute superiority of one approach over the other.

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
Initialization provides an estimate before the iterative optimization in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}
\subsection{The Atropos User Interface}
As with other classes that comprise ANTs, Atropos uses the Insight
Toolkit as a developmental foundation.  This allows us to take
advantage of the mature portions of ITK (e.g. image IO) and ensures
the integrity of the ancillary processes such as those facilitated by
the underlying statistical framework.  Although Atropos is publicly
distributed with the rest of the ANTs package, we plan to contribute
its core elements to the Insight Toolkit where it can be vetted and further
improved by other interested researchers.

An overview of Atropos components can be gleaned, in part, from the flowchart depicted in 
Fig.~\ref{fig:flowchart}.  To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.  
The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.  Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.  Here we  describe these options from the unique perspective of  implementation.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing
\subsubsection{Integrating N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using a method such as the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight Toolkit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.

@@

data with 3\% noise and three levels of bias referred to as 0, 20 and 40 \% RF inhomogeneity.  We

@@

of bias on $($N4 $+$ Atropos$)$ and also the validity of our

@@

Dice overlap, are shown in Figure~\ref{fig:bweb1}. Because overlap

@@

is in Figure~\ref{fig:bwebrf40}.  We supply the information necessary

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.  This is
  sensible in that the prior labels image outcome is sensitive to the
  user choices made.  In this study, we labeled the brain web template
  very sparsely, quickly and crudely and did not expect highly accurate results.  The
  legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4 assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=0.5$, the priors are maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in Figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

@@

We first convert each of the 69 labels within the original evaluation
dataset to an individual image.  The remaining steps,
summarized briefly, are the same for each of the 19 subjects.  We
select one subject as an unlabeled target.  The other 18 datasets are
then mapped (as in the script above) to that subject.  We then deform,
individually, the 69 $\times$ 18 label images to the unlabeled
subject.  The label probability map is gained by averaging the 18
deformed images associated with each label.  We repeat this for each
subject.  The following parameters are the most relevant to this discussion:
{\ttfamily  -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,0.5] 
-m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
Dice overlap, are shown in Figure~\ref{fig:hammers}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small The figure compares the Dice
overlap results from Atropos versus the raw results from majority
voting for each of 68 neuroanatomical regions and, in addition, the
unlabeled portions of the brain from the Hammers evaluation dataset.
We evaluated Atropos via N-fold cross-validation and employed
PriorProbabilityImages for each class where probabilities are gained
by averaging mapped subject labels. The color coding highlights those
regions that have the highest (yellow) and lowest (pink) improvement.
The significance of the improvement, measured by pairwise T-test, is
also shown as is a trinary coding of that improvement as: $+$
significant improvement, $-$ performance reduction, $\sim$ no change.}
\label{fig:hammers}
\end{center}
\end{figure}

@@

\section{Discussion}
We introduced Atropos, the theory and implementation
details and documented its performance in a variety of use cases.  We
also showed evidence that the openly available N4 bias correction can
easily be used with Atropos to improve segmentation.
Furthermore, we used multiple subject BrainWeb data to build dataset
specific priors that provided the most consistent segmentation
performance across tissues.  Finally, we used majority voting to
initialize an Atropos EM solution to a 69 class brain parcellation
problem.  Significant improvements were gained in multiple brain
regions, in particular in temporal lobe cortex, the hippocampi and
amygdalae and the lateral ventricles.  This work, in summary, proves
the applicability of Atropos in both basic and extended use cases.

\subsection{Performance on BrainWeb Data}
Atropos results are competitive with the state of the art.  For
instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
noise BrainWeb single subject data finding 0.932 (GM) and .961 (WM)
Dice. Results on 40\% RF were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
insensitivity to bias similar to our own best results on the 40\% RF, 3\% 
noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice for GM is
0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
results (BrainWeb single 3\% noise) of 0.962 (0\% RF), 0.964 (20\% RF)
and 0.956 (40\%RF) which are slightly higher than either SPM5 or Atropos
results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
for comparison.  Topology preserving methods also perform well.
\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF BrainWeb
single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
respectively.  These are excellent numbers given the additional
constraint applied to the segmentation.  
\cite{Bazin2007c} proposed TOADS and, estimating from the paper's
graph, showed that the average Dice accuracy for 3\% noise various RF
was 0.95--0.96 for WM, 0.93--0.95 for GM and 0.92--0.94 for CSF.
Perhaps the most recent balanced evaluation was performed in
\citep{Klauschen2009}, which reports confusion matrix numbers, rather
than Dice overlap.  Because the absolute true number of GM and WM
voxels for BrainWeb are known, we can convert the confusion matrix to
Dice.  In that case, the SPM5 Dice for BrainWeb single GM and WM is
0.8847 and 0.9085, respectively, while FreeSurfer and FSL's accuracy
is lower. The best GM result for 20 subjects is obtained by SPM5:
0.9295; best WM Dice is from FSL: 0.9499.  We note that
\cite{Klauschen2009} used a comprehensive evaluation where quality of
brain extraction also contributed to the outcome.  Thus, the results
must be interpreted slightly differently than those from other papers.
Finally, we note that in our evaluation of 20 subject BrainWeb data, the prior
probability models performed best of all the models used.  Compared to
the KMeans based segmentation, the prior based segmentation
performance also peaked at lower values of the MRF-$\beta$ term (0.0
and 0.05).  This is reasonable in that the spatial priors themselves
impose a degree of regularity to the segmentation, as in SPM5.

\subsection{Performance on Hammers Data}
Our prior work, \citep{Avants2011}, showed that the majority vote
initialization provided to Atropos by ANTs template mapping is
competitive with \cite{Heckemann2006}.  Overall, the Atropos EM
extension improved these results further.  However, in a few regions
of the mid-brain, the Atropos EM segmentation
performed significantly worse.  This is not surprising, in that
Atropos EM assumes that signal from the likelihood and MRF term is
valuable in improving the segmentation.  This assumption held for
amygdala and lateral ventricles among other areas.  However, in
pallidum and corpus callosum (the most significant areas with loss of
performance), this is not true.  We believe the explanation is that
the intensity varies within these structures and that a more complex
intensity model (or finer parcellation) would be needed here.  An
alternative solution would be to use boundary conditions for these
structures, as in the {\ttfamily PriorLabelImage} Atropos initialization option.
